Loop-Aware Optimizations in PyPy's Tracing JIT

### 2012-08-15 talk/dls2012/paper.tex
describe the scimark benchmakrs

The loop peeling optimization was implemented in the PyPy
framework in about 450 lines of RPython code. That means that the JIT-compilers generated for all
interpreters implemented with RPython now can take advantage of
it. Benchmarks have been executed for a few different interpreters and
we see improvements in several cases. The ideal loop for this optimization
is short and contains numerical calculations with no failing guards and no
external calls. Larger loops involving many operations on complex objects
typically benefit less from it. Loop peeling never makes runtime performance worse, in
the worst case the peeled loop is exactly the same as the preamble. Therefore we
chose to present benchmarks of small numeric kernels where loop peeling can show
its use.

@@

calculations were implemented in both Python and in C and their

@@

\item {\bf sqrt}: approximates the square root of $y$. The approximation is 

@@

  is represented with different type of objects: int's, float's and

@@

\item {\bf conv3}: one-dimensional convolution with fixed kernel-size $3$. A single loop

@@

\item {\bf conv5}: one-dimensional convolution with fixed kernel-size $5$. Similar to conv3, but with 

@@

\item {\bf conv3x3}: two-dimensional convolution with kernel of fixed

@@

\item {\bf dilate3x3}: two-dimensional dilation with kernel of fixed

@@

\item {\bf sobel}: a low-level video processing algorithm used to

@@

The sobel and conv3x3 benchmarks are implemented
on top of a custom two-dimensional array class.
It is
a straightforward implementation providing 2 dimensional
indexing with out of bounds checks. For the C implementations it is
implemented as a C++ class. The other benchmarks are implemented in
plain C. All the benchmarks except sqrt operate on C double-precision floating
point numbers, both in the Python and the C code.

>>>>

The loop peeling optimization was implemented in the PyPy
framework in about 450 lines of RPython code. That means that the JIT-compilers generated for all
interpreters implemented with RPython now can take advantage of
it. Benchmarks have been executed for a few different interpreters and
we see improvements in several cases. The ideal loop for this optimization
is short and contains numerical calculations with no failing guards and no
external calls. Larger loops involving many operations on complex objects
typically benefit less from it. Loop peeling never makes the generated code worse, in
the worst case the peeled loop is exactly the same as the preamble. 
Therefore we
chose to present benchmarks of small numeric kernels where loop peeling can show
its use.

@@

calculations were implemented in both Python, C and Lua and their

@@

\item {\bf sqrt}$\left(T\right)$: approximates the square root of $y$. The approximation is 

@@

  is represented with different type of objects, $T$,: int's, float's and

@@

\item {\bf conv3}$\left(n\right)$: one-dimensional convolution with fixed kernel-size $3$. A single loop

@@

\item {\bf conv5}$\left(n\right)$: one-dimensional convolution with fixed kernel-size $5$. Similar to conv3, but with 

@@

\item {\bf conv3x3}$\left(n\right)$: two-dimensional convolution with kernel of fixed

@@

\item {\bf dilate3x3}$\left(n\right)$: two-dimensional dilation with kernel of fixed

@@

\item {\bf sobel}$\left(n\right)$: a low-level video processing algorithm used to

@@

The sobel and conv3x3 benchmarks are implemented
on top of a custom two-dimensional array class.
It is
a straightforward implementation providing 2 dimensional
indexing with out of bounds checks and 
data stored in row-major order.
For the C implementations it is
implemented as a C++ class. The other benchmarks are implemented in
plain C. All the benchmarks except sqrt operate on C double-precision floating
point numbers, both in the Python and the C code.

In addition we also ported the 
SciMark\footnote{\texttt{http://math.nist.gov/scimark2/}} benchmakts to python, and compared 
their runtimes with the already existing Lua and C implementations. 
This port was performed after the release of the pypy used to run the benchmarks which means that 
these benchmarks have not influenced the pypy implementation.
SciMark consists of

\begin{itemize}
\item {\bf SOR}$\left(n, c\right)$: Jacobi successive over-relaxation on a $n\times n$ grid repreated $c$ times.
The same custom two-dimensional array class as described above is used to represent
the gird.
\item {\bf SparseMatMult}$\left(n, z, c\right)$: Matrix multiplication between a $n\times n$ sparse matrix,
stored in compressed-row format, and a full storage vector, stored in a normal array. The matrix has $z$ non-zero elements and the calculation is repeated $c$ times.
\item {\bf MonteCarlo}$\left(n\right)$: Monte Carlo integration by generating $n$ points uniformly distributed over the unit square and computing the ratio of those within the unit circle.
\item {\bf LU}$\left(n, c\right)$: Computes the LU factorization of a $n \times n$ matrix. The rows of the matrix is shuffled which makes the previously used two-dimensional array class unsuitable. Instead a list of arrays is used to represent the matrix. The calculation is repeated $c$ times.
\item {\bf FFT}$\left(n, c\right)$: Fast Fourier Transform of a vector with $n$ elements, represented as an array, repeated $c$ times.
\end{itemize}
